An elementary and constructive proof of Grothendieck's generic freeness lemma

### 2018-05-15 paper-generic-freeness.tex
Corrected minor typos

\begin{thm}\label{thm:algebraic}Let~$A$ be a reduced ring. Let~$B$ be
an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module.
If~$f = 0$ is the only element of~$A$ such that
\begin{enumerate}
\item the~$A[f^{-1}]$-modules $B[f^{-1}]$ and $M[f^{-1}]$ are free,
\item the~$A[f^{-1}]$-algebra~$B[f^{-1}]$ is of finite presentation, and
\item the~$B[f^{-1}]$-module~$M[f^{-1}]$ is finitely presented,
\end{enumerate}
then~$1 = 0$ in~$A$.
\end{thm}

Previously known proofs either only cover the case where~$A$ is a Noetherian
integral domain, where one can argue by \emph{dévissage} (see for
instance~\cite[Lemme~6.9.2]{ega-4-2},
\cite[Thm.~24.1]{matsumura:commutative-ring-theory}, or
\cite[Thm.~14.4]{eisenbud:commutative-algebra}), or proceed in a series of
intermediate steps, reducing to that case (see for
instance~\cite{staats:generic-freeness} or~\stacksproject{051Q}); but in fact,
a direct proof is possible and shorter.

@@

\begin{thm}\label{thm:geometric}Let~$A$ be a reduced ring. Let~$B$ be
an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module. Then
the space~$\Spec(A)$ contains a dense open~$U$ such that over~$U$,
\begin{enumerate}
\item[(a)] $B^\sim$ and~$M^\sim$ are free as sheaves of~$A^\sim$-modules,
\item[(b)] $B^\sim$ is of finite presentation as a sheaf of~$A^\sim$-algebras, and
\item[(c)] $M^\sim$ is finitely presented as a sheaf of~$B^\sim$-modules.
\end{enumerate}
\end{thm}

Theorem~\ref{thm:geometric} immediately follows from
Theorem~\ref{thm:algebraic} by defining~$U$ as the union of all the basic
opens~$D(f)$ such that~(1),~(2), and~(3) hold. It's clear that~(a),~(b),
and~(c) hold over~$U$, and~$U$ is dense for if~$V$ is an arbitrary open
such that~$U \cap V = \emptyset$, the open~$V$ is itself empty: Let~$h \in A$
such that~$D(h) \subseteq V$. The hypothesis implies the assumptions of
Theorem~\ref{thm:algebraic} for the datum~$(A[h^{-1}], B[h^{-1}], M[h^{-1}])$.
Thus~$1 = 0 \in A[h^{-1}]$, so~$h$ is nilpotent and~$D(h) = \emptyset$.

The new proof was found using a general topos-theoretical technique which we
believe to be useful in other situations as well. This technique allows to view
reduced rings and their modules from a different point of view, one from which
reduced rings look like fields. Since Grothendieck's generic freeness
is trivial for fields, this technique yields a trivial proof for reduced rings.
The proof presented here was obtained by unwinding the topos-theoretic proof,
yielding a self-contained argument without any references to topos theory.
We refer readers who want to learn about this technique to a companion 
paper~\cite{blechschmidt:wlog}.

\textbf{Acknowledgments.} The proof presented here was prompted by
user~HeinrichD on MathOverflow~\cite{mo:kernel} and greatly
benefited from discussions with Martin Brandenburg, who needed the constructive
version in a paper of his~\cite{brandenburg:schur}. I'm grateful to Marc
Nieper-Wißkirchen for carefully guiding my PhD thesis, where \XXX{...}

@@

    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
    \begin{scope}[yshift=4cm, xshift=4cm]
      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=1.0cm,align=center,minimum height=1.0cm}]{
        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
      };
    \end{scope}
  \end{tikzpicture}
  \caption{\label{fig:good}A graphical depiction of a good generating family in
  the special case~$n = 2, m = 1$. The hatched cells indicate vectors which
  have already been removed from the family. The small black squares indicate
  \emph{corners}. If the vector in the red cell will be found to be expressible
  as a linear combination of vectors with smaller index (blue cells), it will
  be removed, along with the vectors in all cells to the top and to the right
  of the red cell.}
\end{figure}

@@

We show that~$(w_j)_{j \in J'}$ is a basis of~$M$ by verifying the assumptions
of Lemma~\ref{lemma:basis}. Thus let~$g \in A$ be given such that one of
the~$w_j$ is an $A[g^{-1}]$-linear combination of generators with smaller index
in~$M[g^{-1}]$. Removing~$w_j = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
vectors~$x_1^{k_1} \cdots x_n^{k_n} v_\ell$ where~$k_1 \geq i_1, \ldots,
k_n \geq i_n$, we obtain a subfamily which is still good for
the~$A[g^{-1}]$-module~$M[g^{-1}]$. By induction, applied
to~$A[g^{-1}]$ and its module~$M[g^{-1}]$, it therefore follows that $A[g^{-1}] = 0$.
This implies that $g = 0$ since~$A$ is reduced.

Similarly, we show that the given good generating family~$(s_i)_{i \in I'}$ is
a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$j =
(\ell,i_1,\ldots,i_n) \in J$, as indicated in Figure~\ref{fig:good},
a way of expressing~$x^{i_1} \cdots x^{i_n} v_\ell = \sum_{(p,k_1,\ldots,k_n)}
a_{j; p,k_1,\ldots,k_n} x^{k_1} \cdots x^{k_n} v_p$ as a linear combination of
generators with strictly smaller index. Then~$M$ is isomorphic to
the~$B$-module
\[ B\langle V_1,\ldots, V_m\rangle/
  (x^{i_1} \cdots x^{i_n} V_\ell - \textstyle\sum_{(p,k_1,\ldots,k_n)}
  a_{j; p,k_1,\ldots,k_n} x_{k_1} \cdots x_{k_n} V_p)_j,
\]
where the index~$j$ ranges over all corners of~$J$, and is therefore finitely
presented. In a similar vain, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
we mod out a suitable ideal with as many generators as corners of~$I$, is
isomorphic to~$B$.

>>>>

\begin{thm}\label{thm:algebraic}Let~$A$ be a reduced ring. Let~$B$ be
an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module.
If~$f = 0$ is the only element of~$A$ such that
\begin{enumerate}
\item the~$A[f^{-1}]$-modules $B[f^{-1}]$ and $M[f^{-1}]$ are free,
\item the~$A[f^{-1}]$-algebra~$B[f^{-1}]$ is of finite presentation and
\item the~$B[f^{-1}]$-module~$M[f^{-1}]$ is finitely presented,
\end{enumerate}
then~$1 = 0$ in~$A$.
\end{thm}

Previously known proofs either only cover the case where~$A$ is a Noetherian
integral domain, where one can argue by \emph{dévissage} (see for
instance~\cite[Lemme~6.9.2]{ega-4-2},
\cite[Thm.~24.1]{matsumura:commutative-ring-theory} or
\cite[Thm.~14.4]{eisenbud:commutative-algebra}), or proceed in a series of
intermediate steps, reducing to that case (see for
instance~\cite{staats:generic-freeness} or~\stacksproject{051Q}); but in fact,
a direct proof is possible and shorter.

@@

\begin{thm}\label{thm:geometric}Let~$A$ be a reduced ring. Let~$B$ be
an~$A$-algebra of finite type. Let~$M$ be a finitely generated~$B$-module. Then
the space~$\Spec(A)$ contains a dense open~$U$ such that over~$U$,
\begin{enumerate}
\item[(a)] $B^\sim$ and~$M^\sim$ are free as sheaves of~$A^\sim$-modules,
\item[(b)] $B^\sim$ is of finite presentation as a sheaf of~$A^\sim$-algebras and
\item[(c)] $M^\sim$ is finitely presented as a sheaf of~$B^\sim$-modules.
\end{enumerate}
\end{thm}

Theorem~\ref{thm:geometric} immediately follows from
Theorem~\ref{thm:algebraic} by defining~$U$ as the union of all those basic
opens~$D(f)$ such that~(1),~(2) and~(3) hold. It's clear that~(a),~(b)
and~(c) hold over~$U$, and~$U$ is dense for if~$V$ is an arbitrary open
such that~$U \cap V = \emptyset$, the open~$V$ is itself empty: Let~$h \in A$ be
such that~$D(h) \subseteq V$. The hypothesis implies the assumptions of
Theorem~\ref{thm:algebraic} for the datum~$(A[h^{-1}], B[h^{-1}], M[h^{-1}])$.
Thus~$1 = 0 \in A[h^{-1}]$, so~$h$ is nilpotent and~$D(h) = \emptyset$.

The new proof was found using a general topos-theoretical technique which we
believe to be useful in other situations as well. This technique allows to view
reduced rings and their modules from a different point of view, one from which
reduced rings look like fields. Since Grothendieck's generic freeness
is trivial for fields, this technique yields a trivial proof for reduced rings.
The proof presented here was obtained by unwinding the topos-theoretic proof,
yielding a self-contained argument without any references to topos theory.
We refer readers who want to learn about this technique to a forthcoming companion 
paper~\cite{blechschmidt:wlog}.

\textbf{Acknowledgments.} The proof presented here was prompted by
user~HeinrichD on MathOverflow~\cite{mo:kernel} and greatly
benefited from discussions with Martin Brandenburg, who employed the constructive
version in a paper of his~\cite{brandenburg:schur}. I'm grateful to Marc
Nieper-Wißkirchen for carefully guiding my PhD studies at the University of
Augsburg, where most of the work for this paper was carried out.

@@

    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
    \begin{scope}[yshift=4cm, xshift=4cm]
      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=1.0cm,align=center,minimum height=1.0cm}]{
        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
      };
    \end{scope}
  \end{tikzpicture}
  \caption{\label{fig:good}A graphical depiction of a good generating family (the non-hatched cells) in
  the special case~$n = 2, m = 1$, writing~``$x$'' and~``$y$'' for~$x_1$ and~$x_2$. The hatched cells indicate vectors which
  have already been removed from the family. The small black squares indicate
  \emph{corners}. If the vector in the red cell will be found to be expressible
  as a linear combination of vectors with smaller index (blue cells), it will
  be removed, along with the vectors in all cells to the top and to the right
  of the red cell.}
\end{figure}

@@

We show that~$(w_j)_{j \in J'}$ is a basis of~$M$ by verifying the assumptions
of Lemma~\ref{lemma:basis}. Thus let~$g \in A$ be given such that one of
the~$w_j$ is an $A[g^{-1}]$-linear combination of generators with smaller index
in~$M[g^{-1}]$. Removing~$w_j = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
vectors~$x_1^{k_1} \cdots x_n^{k_n} v_\ell$ where~$k_1 \geq i_1, \ldots,
k_n \geq i_n$, we obtain a subfamily which is still good for
the~$A[g^{-1}]$-module~$M[g^{-1}]$. By the induction hypothesis, applied
to~$A[g^{-1}]$ and its module~$M[g^{-1}]$, it therefore follows that $A[g^{-1}] = 0$.
This implies that $g = 0$ since~$A$ is reduced.

Similarly, we show that the given good generating family~$(s_i)_{i \in I'}$ is
a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$j =
(\ell,i_1,\ldots,i_n) \in J$, as indicated in Figure~\ref{fig:good},
a way of expressing~$x^{i_1} \cdots x^{i_n} v_\ell = \sum_{(p,k_1,\ldots,k_n)}
a_{j; p,k_1,\ldots,k_n} x^{k_1} \cdots x^{k_n} v_p$ as a linear combination of
generators with strictly smaller index. Then~$M$ is isomorphic to
the~$B$-module
\[ B\langle V_1,\ldots, V_m\rangle/
  (x^{i_1} \cdots x^{i_n} V_\ell - \textstyle\sum_{(p,k_1,\ldots,k_n)}
  a_{j; p,k_1,\ldots,k_n} x_{k_1} \cdots x_{k_n} V_p)_j,
\]
where the index~$j$ ranges over all corners of~$J$, and is therefore finitely
presented. In a similar vain, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
we mod out a suitable ideal with as many generators as corners of~$I'$, is
isomorphic to~$B$.
